\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2017-10-26 document/XSGRB_sample.tex
Update 26/10
More intro and conclusions
AA changed to nm

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs (typically
R=$\lambda/\Delta\lambda$$<$1000) \citep[e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability
to secure information for even very faint targets. This allows the measurements
of a number of important parameters such as redshifts and spectral slopes. For
a handful of very bright afterglows high-resolution (typically R$>$20000)
spectra were secured and for these events much more information about
conditions inside the host galaxies were extracted 
\citep[e.g.,][]{Fiore,Thoene07,Prochaska07,Vreejswijk07,CastroTirado2010}.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd
generation instruments at the ESO Very Large Telescope (VLT). It is a
high-throughput echelle spectrograph covering the full spectra range from the
UV atmospheric cutoff to the near-infrared (near-IR) K-band with three spectral
arms (UVB, VIS, NIR). It was designed very much with transient source follow-up
in mind as the fading luminosities of such sources makes in urgent to secure as
extensive coverage as possible in the shortest possible time. At the same time
the resolution was set to be in the range 4000--9000 in order to be able to get
a large useful spectra range between the many sky-background emission lines in
the red and near-IR spectral ranges. In this paper we present the results
of a dedicated effort over the years 2009 -- 2016 to use the X-shooter
spectrograph to secure spectroscopic observations of
afterglows and host galaxies of GRBs detected by {\it Swift}. 
All data resulting from the survey are publicly availble in reduced form.

@@

X-shooter can cover the spectral wavelength region from 3000 \AA~to 24800
\AA~in a single exposure, by splitting the light in three separate
spectroscopic arms using dichroics.  This way each arm work as a separate
instrument, each functioning as its own echelle spectrograph.  The ultraviolet
blue(UVB) arm covering 3000 - 5500 \AA, the visual(VIS) arm covering 5500 -
10200 \AA, and the near-infrared()NIR) arm covering 10200 - 24800 \AA~ with the
possibility of applying a k-band blocking filter cutting the coverage of the
NIR arm at 10200 - 21000 \AA.
\\

@@

The image is rectified onto an equidistant grid with a dispersion sampling of
0.2 \AA/pixel and a 0.16 "/pixel spatial sampling for the UVB and VIS arm and
0.6 \AA/pixel with a 0.21 "/pixel in the NIR arm.  Because the tabulated
resolution is a lower limit, by choosing a sampling of 0.2 \AA/pixel, we ensure
that the bluest part of neither of the arms have a sampling lower than the
Nyquist sampling rate of 2 pixels per resolution FWHM.

@@

The resulting offsets, which were smaller than 0.1 \AA~in the UVB and VIS data
and smaller than 0.5 \AA~in the NIR spectra, but changing over short period of
time were applied to the corresponding spectra.

@@

The sample we present here include spectroscopic observations of 86 
systems fulfilling our sample criteria of which 72 are afterglow spectra 
and 14 are observations of the underlying host galaxies. 
UPDATE THIS.

Our sample serves the purpose to further characterize the environments 
of GRBs that was also much advanced by the previous surveys based primarily on 
lower-resolution spectographs. The most striking result found is that GRBs
typically are located behind very large column densities of neutral hydrogen.
Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ quasars have a handful such systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep[e.g., five in][]{Noterdaeme2012}. Long GRB afterglow spectra, by contrast, 
reveal such
systems in the majority of cases \citep[][and this work]{Jakobsson2006, Fynbo2009}. 
With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, etc.

It seems important to end this paper by stressing that the potential of 
using GRB sightlines as probes is far from fully harvested. The sample of
sightlines probed by our spectra is not representative for all GRB 
sightlines as we have shown (CONFIRM) and consistent with earlier findings
\citep[e.g.,][]{Fynbo2009}. \cite{Kruhler2013} convincingly argue, that 
very rich sightlines like that probed by the remarkable GRB\,080607 
\citep{Prochaska09} are probably significantly more frequent than in the sightlines 
sampled by our spectra. However, with current instrumentation these sightlines are
out of reach except under very fortunate circumstances as in the case of 
GRB\,080607 when the afterglow could be observed only a few minutes after the 
burst with a 10-m class telescope. Observations of such sightlines with 
X-shooter-like spectrographs on th enext generation of 20--40-m telescopes
is likely to be very rewarding.

@@

\subsection{GRB120224A (z = 1.10 NEW!!!)}
The data presented here has formed the basis of GCN \#
12991\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/12991.gcn3}}, and has also
been published in \citet{Kruhler2015}. Starting 19.8 hours after the GRB
trigger, a total exposure time of 2.4 ks reveals a faint continuum, starting at
$\sim$ 7000 \AA~and extending all the way through 25000 \AA. We detect a $\sim
2 \sigma$ emission line which, if interpreted as \ha, gives $z = 1.10$,
supporting the redshift reported by \citet{Kruhler2015}.

@@

\subsection{GRB120712A (z = 4.175)}
The data presented here also formed the basis of GCN \#
13460\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13460.gcn3}} and is not
published elsewhere. 4.8 ks integration time, starting 10.5 hours after the BAT
trigger, shows a bright afterglow continuum starting at $\sim$ 4720 \AA,
signifying the onset of the Lyman alpha forest, for a GRB located at $z =
4.175$. Absorption features from \lya, \feii, \mgii~and \SIii~are readily
detected at a consistent redshift.

@@

\subsection{GRB120805A (z $\sim$ 3.9 NEW!!!)}
A separate reduction of this burst has been published in \citet{Kruhler2015},
but not otherwise. Starting 9 days after the burst trigger, this is host
observation and does not contain any afterglow continuum. In 3.6 ks integration
time, we detect a faint continuum visible from 4500 \AA~and all the way through
21000 \AA, in contrast to what is found previously. The continuum from 4500 -
6000 \AA is detected at very low significance. If the drop at 4500 \AA~is the
Lyman limit, this fits with Lyman alpha at $\sim$ 6000 \AA, giving $z \sim
3.9$. The absence of nebular lines if due to \oii~falling in a telluric
absorption band and the rest being shifted out of the wavelength coverage.

@@

\subsection{GRB120909A (z = 3.929)}
The data presented here has formed the basis of GCN \#
13730\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13730.gcn3}}, but is not
published otherwise. A very rapid follow-up, starting only 1.7 hours after the
BAT trigger, this 1.2 ks observation captures a very bright afterglow
continuum, starting at 4500 \AA, signifying the onset of the Lyman limit for a 
system at $z = 3.929$. Absorption from high-column density hydrogen leaves very
prominent absorption features in the form of \lya, \lyb, and \lyg, visible in
the Lyman alpha forest. Metal absorption lines arising from \feii, \NIii,
\SIii, \sii, \alii, \aliii, \cii, \oi, \civ, and \znii~are all detected along
with the corresponding fine-structure lines from (\feii$^*$, \SIii$^*$,
\oi$^*$, \oi$^**$, \cii$^*$), securely anchoring the redshift of the host.

@@

\subsection{GRB161007A (z =4.6 ??? NEW!!!)}
This data has not been published elsewhere. Observations for GRB161007A started
323 hours after the burst trigger and contains the potential host. 4 x 600 seconds
of observations reveals a faint continuum rising abruptly above the noise at
$\sim$ 6850 \AA~and continuing through 21000 \AA. A very low significance
continuum is detected at shorter wavelengths, down to $\sim$ 6000 \AA. If the
host is located at $z \sim 4.6$, the drop in continuum flux is the Lyman alpha
break and the absence of nebular emission lines is due to \oii being shifted
out of the wavelength coverage. Alternatively, an early-type host at $z = 0.71$
could exhibit the 4000 \AA break at 6000 \AA, but due to the preference of
long-duration GRBs for star-forming galaxies, this is the least likely
explanation, why we believe the high-z solution.

\subsection{GRB161014A (z =2.823)}
The data presented here also formed the basis of GCN \#
20061\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/20061.gcn3}}, but is not
published elsewhere. Starting 11.6 hours after the GRB trigger, 4.8 ks of
integration time captures the afterglow continuum across all three
spectroscopic arms. A broad absorption trough due to Lyman alpha is visible,
along with metal absorption features from \mgii, \SIii, \cii, \civ, \alii,
\aliii, and	\feii, all at $z =2.823$. Similar to GRB140506 \citep{Fynbo2014}, a
break in the continuum shape is detected bluewards of 6000\AA, possible
signifying some anomalous form of extinction.

>>>>

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs 
\citep[typically
R=$\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability
to secure information for even very faint targets. This allows the measurements
of a number of important parameters such as redshifts, spectral slopes,
and extinction. For
a handful of very bright afterglows high-resolution (typically R$>$20000)
spectra were secured and for these events much more information about
conditions inside the host galaxies were extracted 
\citep[e.g.,][]{Fiore,Thoene07,Prochaska07,Vreejswijk07,CastroTirado2010}.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd
generation instruments at the ESO Very Large Telescope (VLT). It is a
high-throughput echelle spectrograph covering the full spectral range from the
UV atmospheric cutoff around 300 nm to the near-infrared (near-IR) K-band 
ending around 2500 nm with three spectral
arms (UVB, VIS, NIR). It was designed very much with transient source follow-up
in mind as the fading luminosities of such sources makes in urgent to secure as
extensive coverage as possible in the shortest possible time. At the same time
the resolution was set to be in the range 4000--9000 in order to be able to get
a large useful spectral range between the many sky-background emission lines in
the red and near-IR spectral ranges. In this paper we present the results
of a dedicated effort over the years 2009 -- 2016 to use the X-shooter
spectrograph to secure spectroscopic observations of
afterglows and host galaxies of GRBs detected by {\it Swift}. 
All data resulting from the survey are publicly availble in reduced form.

@@

X-shooter can cover the spectral wavelength region from 300 nm to 2480
nm in a single exposure, by splitting the light in three separate
spectroscopic arms using dichroics.  This way each arm work as a separate
instrument, each functioning as its own echelle spectrograph.  The ultraviolet
blue(UVB) arm covering 3000 - 550 nm, the visual(VIS) arm covering 550 -
1020 nm, and the near-infrared()NIR) arm covering 10200 - 2480 nm with the
possibility of applying a k-band blocking filter cutting the coverage of the
NIR arm at 1020 - 2100 nm.
\\

@@

The image is rectified onto an equidistant grid with a dispersion sampling of
0.02 nm/pixel and a 0.16 "/pixel spatial sampling for the UVB and VIS arm and
0.06 nm/pixel with a 0.21 "/pixel in the NIR arm.  Because the tabulated
resolution is a lower limit, by choosing a sampling of 0.02 nm/pixel, we ensure
that the bluest part of neither of the arms have a sampling lower than the
Nyquist sampling rate of 2 pixels per resolution FWHM.

@@

The resulting offsets, which were smaller than 0.01 nm in the UVB and VIS data
and smaller than 0.05 nm in the NIR spectra, but changing over short period of
time were applied to the corresponding spectra.

@@

The sample we present here include spectroscopic observations of 86 
systems fulfilling our sample criteria of which 72 are afterglow spectra 
and 14 are observations of the underlying host galaxies. 
UPDATE THIS.
All spectra have been made publicly available in the reduced form used
in this paper.

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectographs. The most striking result found is that GRBs
typically are located behind very large column densities of neutral hydrogen.
Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ quasars have a handful such systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified \citep[e.g., five
in][]{Noterdaeme2012}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep[][and this work]{Jakobsson2006,
Fynbo2009,Cucchiara2015}.  With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we hence characterize the properties of
star-forming galaxies over cosmic history in terms of redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, extinction curves,
etc.  A number of independent papers have been published or submitted for
publication focusing on many of these specific issues of our sample such as
extinction curves [][see also
\cite{Fynbo2014,Heintz2017}]\citep{Japelj2015,Zafar17}, emission lines from the
underlying host galaxies \citep{Kruhler2015}, the frequency of intervening
\ion{Mg}{ii} absorbers \citep{Christensen2017}, and escape of ionizing
radiation \citep{Tanvir17}.

It seems important to end this paper by stressing that the potential of 
using GRB sightlines as probes is far from fully harvested. The sample of
sightlines probed by our spectra is not representative for all GRB 
sightlines as we have shown and consistent with earlier findings
from samples based on low-resolution spectroscopy \citep[e.g.,][]{Fynbo2009}
and from studies of complete samples of GRB host galaxies 
\citep{Hjorth2012,Covinto2013,Perley2016}. 
\cite{Kruhler2013} convincingly argue, that 
very rich sightlines like that probed by the remarkable GRB\,080607 
\citep{Prochaska09} are probably significantly more frequent than in the sightlines 
sampled by our spectra. However, with current instrumentation these sightlines are
out of reach except under very fortunate circumstances as in the case of 
GRB\,080607 when the afterglow could be observed only a few minutes after the 
burst with a 10-m class telescope. Observations of such sightlines with 
X-shooter-like spectrographs on th enext generation of 20--40-m telescopes
is likely to be very rewarding.

@@

\subsection{GRB120224A (z = 1.10 NEW!!!)}
The data presented here has formed the basis of GCN \#
12991\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/12991.gcn3}}, and has also
been published in \citet{Kruhler2015}. Starting 19.8 hours after the GRB
trigger, a total exposure time of 2.4 ks reveals a faint continuum, starting at
$\sim$ 700 nm~and extending all the way through 2500 nm. We detect a $\sim
2 \sigma$ emission line which, if interpreted as \ha, gives $z = 1.10$,
supporting the redshift reported by \citet{Kruhler2015}.

@@

\subsection{GRB120712A (z = 4.175)}
The data presented here also formed the basis of GCN \#
13460\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13460.gcn3}} and is not
published elsewhere. 4.8 ks integration time, starting 10.5 hours after the BAT
trigger, shows a bright afterglow continuum starting at $\sim$ 472 nm,
signifying the onset of the Lyman alpha forest, for a GRB located at $z =
4.175$. Absorption features from \lya, \feii, \mgii~and \SIii~are readily
detected at a consistent redshift.

@@

\subsection{GRB120805A (z $\sim$ 3.9 NEW!!!)}
A separate reduction of this burst has been published in \citet{Kruhler2015},
but not otherwise. Starting 9 days after the burst trigger, this is host
observation and does not contain any afterglow continuum. In 3.6 ks integration
time, we detect a faint continuum visible from 450 nm and all the way through
2100 nm, in contrast to what is found previously. The continuum from 4500 -
600 nm is detected at very low significance. If the drop at 450 nm is the
Lyman limit, this fits with Lyman alpha at $\sim$ 600 nm, giving $z \sim
3.9$. The absence of nebular lines if due to \oii~falling in a telluric
absorption band and the rest being shifted out of the wavelength coverage.

@@

\subsection{GRB120909A (z = 3.929)}
The data presented here has formed the basis of GCN \#
13730\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13730.gcn3}}, but is not
published otherwise. A very rapid follow-up, starting only 1.7 hours after the
BAT trigger, this 1.2 ks observation captures a very bright afterglow
continuum, starting at 450 nm, signifying the onset of the Lyman limit for a 
system at $z = 3.929$. Absorption from high-column density hydrogen leaves very
prominent absorption features in the form of \lya, \lyb, and \lyg, visible in
the Lyman alpha forest. Metal absorption lines arising from \feii, \NIii,
\SIii, \sii, \alii, \aliii, \cii, \oi, \civ, and \znii~are all detected along
with the corresponding fine-structure lines from (\feii$^*$, \SIii$^*$,
\oi$^*$, \oi$^**$, \cii$^*$), securely anchoring the redshift of the host.

@@

\subsection{GRB161007A (z =4.6 ??? NEW!!!)}
This data has not been published elsewhere. Observations for GRB161007A started
323 hours after the burst trigger and contains the potential host. 4 x 600 seconds
of observations reveals a faint continuum rising abruptly above the noise at
$\sim$ 685 nm and continuing through 2100 nm. A very low significance
continuum is detected at shorter wavelengths, down to $\sim$ 600 nm. If the
host is located at $z \sim 4.6$, the drop in continuum flux is the Lyman alpha
break and the absence of nebular emission lines is due to \oii being shifted
out of the wavelength coverage. Alternatively, an early-type host at $z = 0.71$
could exhibit the 400 nm break at 600 nm, but due to the preference of
long-duration GRBs for star-forming galaxies, this is the least likely
explanation, why we believe the high-z solution.

\subsection{GRB161014A (z =2.823)}
The data presented here also formed the basis of GCN \#
20061\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/20061.gcn3}}, but is not
published elsewhere. Starting 11.6 hours after the GRB trigger, 4.8 ks of
integration time captures the afterglow continuum across all three
spectroscopic arms. A broad absorption trough due to Lyman alpha is visible,
along with metal absorption features from \mgii, \SIii, \cii, \civ, \alii,
\aliii, and	\feii, all at $z =2.823$. Similar to GRB140506 \citep{Fynbo2014}, a
break in the continuum shape is detected bluewards of 600 nm, possible
signifying some anomalous form of extinction.
